Formal Verification Framework Based on Symbolic Execution for Smart Contract

### 2023-12-15 papers/static-checker/ecoop2024/ecoop2024.tex
write the static checker part

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
code := {c}
input := t
output := t'
pre-condition := cd
post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the syntax of our property specification language. Here, \KMCONTRACT, \KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KNAME\ is replaced by the contract name. In the input and output fields, the input and output patterns are specified. Since Michelson operates on a stack, we assume that the stack has only one element, as specified in the input pattern. Similarly, the output stack is assumed to be a stack with only one element, which is specified in the output field. As we have mentioned before, the output should be a pair consisting of an operation list and a storage. In our syntax, for verification purpose, we omit the operation list and then the output is represented by the patern of the value in the storage after the execution is finished. The input and output patterns represent Michelson terms, which can be either concrete values or symbolic values, including variables. When a variable appears for the first time, it needs to be accompanied by its corresponding type declaration. However, subsequent mentions can omit the type declaration. Formally, the pre and post-condition instances are formulas in first-order logic, and they can be handled by an SMT solver, namely the Z3 solver in our framework. The detail of the language grammar is described in in Figure \ref{}. Formulas that are used in assertions or added to solvers are terms of Boolean sort. Otherwise, terms of Boolean and non-Boolean sort may be mixed in any combination where sorts match up. Universal and existential quantifiers bind variables to the scope of the quantified formula.

Let's consider the following example of the property specification for the contract named Add:
\begin{lstlisting}[float,captionpos=b,caption={The specification of Add contracr},label={lst:safer-contract-module},numbers=left]
mcontract Add = spec
code := {UNPAIR; COMPARE; GE; 
        IF {ADD}{PUSH string 'Unexpected Pair'; FAILWITH}}
input := Pair (x : int) (y : int)
output := (z : int)
pre-condition := (x >= 5) | (y >= 3)
post-condition := (z >= 8)
\end{lstlisting}
Here, the program takes as input a stack with only one element, which is a pair of two integers, symbolically specified in the Input field. It then unpairs the input pair to obtain two elements as two integers \lstinline/x/ and \lstinline/y/.  It then proceeds to destructure the input pair, yielding two integers. Subsequently, the program performs a comparison between these two integers. If  \lstinline/x/ is greater than or equal to \lstinline/y/, the program constructs a new stack with a solitary element at the top representing the sum of  \lstinline/x/ and \lstinline/y/, designated as \lstinline/z/. Otherwise, the execution terminates with an error message containing the content 'Unexpected pair.' The specification continues by defining pre and post conditions, where the precondition states the assumption that the integer  \lstinline/x/ is greater than or equal to 5 and \lstinline/y/ is greater than or equal to 3. Here, the symbol  \lstinline/|/ is the conjuration. The postcondition asserts that the result sum \lstinline/z/ is greater than or equal to 8.

@@

To illustrate this, let's consider the following auction smart contract. This contract has two entrypoints, bid and close, which serve as functions for bidding and closing the contract, respectively. The smart contract storage on the blockchain is a pair containing the first element as a boolean value that indicates whether the contract is still open for bidding or already closed. The second element is another nested pair, where the first element  stores the highest bidder's address, and the second element is the contract owner's address. The current highest bid is reflected in the balance of the contract.

@@

\subsection{User specification}
In the user specification phase, users compose the specification in our domain-specific language using the syntax outlined in Section \ref{sec:domain-specific-language}. Note that Michelson code, whether authored by users, obtained from the Tezos blockchain, or compiled from higher-level languages, can be seamlessly employed within our framework. The specification is subsequently parsed into an Abstract Syntax Tree (AST) in OCaml. The code and the input pattern are then utilized as inputs for the interpreter during the intermediate phase to facilitate symbolic execution. In the next phase, the pre and post predicates are transformed into Z3 formulas by the Z3 converter tool.
\subsection{Michelson interpreter and Z3 converter}
The interpreter implements the execution model described in \ref{sec:symbolic-execution-model}. When provided with a concrete input, the interpreter produces a concrete value. In the presence of symbolic input, the result comprises a set of symbolic states represented as terms in OCaml. Due to Michelson being built in OCaml, the conversion of Michelson types or terms to their OCaml counterparts is straightforward. Each state contains the output stack, which has only one element consisting of an operation list and storage, along with the path conditions represented as predicates in conjunctive form. This output is then fed into the Z3 converter, which handles two different types of terms: assertion terms from the AST representing the pre and post-condition specifications, and terms resulting from the interpreter. We utilize the Z3 library in OCaml to construct this converter. Some adjustments are necessary to convert these terms into Z3 formulas. For example, as there is no natural number type (\TNAT) in Z3, the term is converted as an integer type, accompanied by constraints asserting that it is greater than or equal to zero. Several constants, such as \CAMOUNT, \CBALANCE, \CSENDER\ and cryptographic functions, such as \FSHA, \FHASHKEY\ are predefined in the environment that works for any smart contract specifications.
\subsection{Static checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions. The static checker performs three distinct features: the output check, fail condition and property check.

\subsubsection{Output check} First is the output check, where the checker verifies whether the output pattern matches. It basiclly matches the output pattern with the correcspoding part in the final states resulting from the symbolic intepreter.  When the input pattern is a concrete value, the checker can be used as a test to verify if the output aligns with the expectations specified in the output field.

\subsubsection{Fail condition}

>>>>

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
code := {c}
input := t
output := t'
pre-condition := cd
post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the syntax of our property specification language. Here, \KMCONTRACT, \KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KNAME\ is replaced by the contract name. In the input and output fields, the input and output patterns are specified. Since Michelson operates on a stack, we assume that the stack has only one element, as specified in the input pattern. Similarly, the output stack is assumed to be a stack with only one element, which is specified in the output field. As we have mentioned before, the output should be a pair consisting of an operation list and a storage. In our syntax, for verification purpose, we omit the operation list and then the output is represented by the patern of the value in the storage after the execution is finished. The input and output patterns represent Michelson terms, which can be either concrete values or symbolic values, including variables. When a variable appears for the first time, it needs to be accompanied by its corresponding type declaration. However, subsequent mentions can omit the type declaration. Formally, the pre and post-condition instances are formulas in first-order logic, and they can be handled by an SMT solver, namely the Z3 solver in our framework. The detail of the language grammar is described in in Figure \ref{}. Formulas that are used in assertions or added to solvers are terms of Boolean sort. Otherwise, terms of Boolean and non-Boolean sort may be mixed in any combination where sorts match up. Universal and existential quantifiers bind variables to the scope of the quantified formula.
\begin{lstlisting}[float,captionpos=b,caption={The specification of Add contracr},label={lst:add-contract-specification},numbers=left]
mcontract Add = spec
code := {UNPAIR; COMPARE; GE; 
        IF {ADD}{PUSH string 'Unexpected Pair'; FAILWITH}}
input := Pair (x : int) (y : int)
output := (z : int)
pre-condition := (x >= 5) | (y >= 3)
post-condition := (z >= 8)
\end{lstlisting}
Let's consider the  example of the property specification for the contract named Add as in Listing \ref{lst:add-contract-specification}. Here, the program takes as input a stack with only one element, which is a pair of two integers, symbolically specified in the Input field. It then unpairs the input pair to obtain two elements as two integers \lstinline/x/ and \lstinline/y/.  It then proceeds to destructure the input pair, yielding two integers. Subsequently, the program performs a comparison between these two integers. If  \lstinline/x/ is greater than or equal to \lstinline/y/, the program constructs a new stack with a solitary element at the top representing the sum of  \lstinline/x/ and \lstinline/y/, designated as \lstinline/z/. Otherwise, the execution terminates with an error message containing the content 'Unexpected pair.' The specification continues by defining pre and post conditions, where the precondition states the assumption that the integer  \lstinline/x/ is greater than or equal to 5 and \lstinline/y/ is greater than or equal to 3. Here, the symbol  \lstinline/|/ is the conjuration. The postcondition asserts that the result sum \lstinline/z/ is greater than or equal to 8.

@@

To illustrate this, let's consider the  auction smart contract serving the auction events. The contract has two entrypoints, bid and close, which serve as functions for bidding and closing the contract, respectively. The smart contract storage on the blockchain is a pair containing the first element as a boolean value that indicates whether the contract is still open for bidding or already closed. The second element is another nested pair, where the first element  stores the highest bidder's address, and the second element is the contract owner's address. The current highest bid is reflected in the balance of the contract.

@@

\subsection{User specification}
In the user specification phase, users compose the specification in our domain-specific language using the syntax outlined in Section \ref{sec:domain-specific-language}. Note that Michelson code, whether authored by users, obtained from the Tezos blockchain, or compiled from higher-level languages, can be seamlessly employed within our framework. The specification is subsequently parsed into an Abstract Syntax Tree (AST) in OCaml. The code and the input pattern are then utilized as inputs for the interpreter during the intermediate phase to facilitate symbolic execution. In the next phase, the pre and post predicates are transformed into Z3 formulas by the Z3 converter tool.
\subsection{Michelson interpreter and Z3 converter}
The interpreter implements the execution model described in \ref{sec:symbolic-execution-model}. When provided with a concrete input, the interpreter produces a concrete value. In the presence of symbolic input, the result comprises a set of symbolic states represented as terms in OCaml. Due to Michelson being built in OCaml, the conversion of Michelson types or terms to their OCaml counterparts is straightforward. Each state contains the output stack, which has only one element consisting of an operation list and storage, along with the path conditions represented as predicates in conjunctive form. This output is then fed into the Z3 converter, which handles two different types of terms: assertion terms from the AST representing the pre and post-condition specifications, and terms resulting from the interpreter. We utilize the Z3 library in OCaml to construct this converter. Some adjustments are necessary to convert these terms into Z3 formulas. For example, as there is no natural number type (\TNAT) in Z3, the term is converted as an integer type, accompanied by constraints asserting that it is greater than or equal to zero. Several constants, such as \CAMOUNT, \CBALANCE, \CSENDER\ and cryptographic functions, such as \FSHA, \FHASHKEY\ are predefined in the environment that works for any smart contract specifications.
\subsection{Static checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions.

What the tool does is first build the input state from the input pattern, runs the symbolic interpreter on the input stack with the provided code to generate all symbolic execution states. It's important to note that all states are reachable from the initial state, as any unreachable states are discarded from the final state spaces using the Z3 solver to check the path condition predicate. Using the output pattern and the pre- and post-conditions converted by the Z3 conveter, the tool then performs different checks. The static checker offers three distinct features: output check, fail condition, and property check.

\subsubsection{Output check} 
The first check is the output check, where the checker verifies whether the output pattern matches. The final state contains a stack with only one element as \Stack\ \STACKCONCAT\ \EMPTYSTACK. There are two different forms of the final state. In the case where the code is given as incomplete smart contract code, the element \Stack\ contains just a normal term and its corresponding type as (\Term, \TY). However, in the case of a complete smart contract code, it has the form (\PAIR\ \VOPERATIONLIST\ \VSTORAGE, \TPAIR\ (\TOPERATIONLIST) \TYS) \STACKCONCAT\ \EMPTYSTACK, where \VSTORAGE\ is the term represting the storage and \VOPERATIONLIST\ repesents a operation list. Given an output pattern $ \tau$, the checker scans all the final states, which does not contain a failwith stack and checks whether the output pattern matches. In the former case, it matches $ \tau$  with the term \Term. In the latter case, it essentially matches with the corresponding part \VSTORAGE\ in the final states resulting from the symbolic interpreter. When the input pattern is a concrete value, the result from the interpreter is also concrete, and then the checker can be used as a test to verify if the output aligns with the expectations specified in the output field.

Lets get back to the example in Listing \ref{lst:add-contract-specification}. This is an example of incomplete smart contract code. The specification states that the output of the code must be a stack that contains only one element of type \TINT\ as symbolicly representing as a variable \VZ\ of type \TINT.  In this example, there are two final states, with the first one being a failwith state recording the failwith condition when the first element of the pair is less than the second element. The failwith state contains the message 'Unexpected Pair,' describing the reason for the failure. The second state is a reachable state that contains, at the top element, an integer value, which is the sum of the two elements of the pair. The tool results in \TRUE\ for the output check for this example, as expected by the code. However, this test could fail if, for example, the output pattern is incorrectly specified as a variable of type \TSTRING, \TMUTEZ, or even \TNAT, as the types wouldn't match with the actual result. If the input pattern is concrete, for instance, a pair of numbers 5 and 3 as Pair 5 3, then the expected output should be the number 8. This can be verified if the output pattern is also specified as a concrete value.

The second example in Listing \ref{lst:auction-contract-specification} is the specification for a smart contract code. The output pattern represents the updated storage after executing the code. There are several final states resulting from the symbolic execution. The failwith state is ignored for the output check, and all other reachable states are matched with the output term. The result is \TRUE\ for this example.

\subsubsection{Fail condition}
A Michelson contract can fail due to a programmer-induced condition caused by the instruction \FAILWITH. It terminates contract execution with an error message reported back to the caller. This error message includes the top value on the stack. These failures are important in smart contract programming to guarantee that a smart contract call should not be successful under specific circumstances. Smart contract calls are expected to fail due to malicious behavior, such as an unauthorized caller or invalid arguments passed along the call. For example, the auction contract should not be closed by someone who is not the owner of the contract, and the bidding should not succeed if the bid is not strictly higher than the current highest bid. Several smart contract standards, such as , require the contract to fail with specific messages under some conditions. Therefore, it is essential to check the fail conditions of a contract.

The static checker provides a feature to generate all fail conditions implemented in a smart contract code. The tool searches for all failwith states, resulting from an execution terminated by the \FAILWITH\ instruction, from the final state space, collects the path conditions and the recorded message. The result returned to users includes the fail condition together with the message. For example, for the specification in Listing \ref{lst:add-contract-specification}, the fail condition is recorded as \lstinline/(x < y): 'Unexpected Pair'/. For the auction contract, there are several fail conditions. For the entrypoint \lstinline/bid/, the fail condition is as shown in Listing \ref{lst:fail-condition-auction-contract-bid} and for the \lstinline/close/ entrypoint, it is recorded in Listing \ref{}.

\begin{lstlisting}[float,captionpos=b,caption={Auction contract specification},label={lst:fail-condition-auction-contract},numbers=left]
1. not (auction_open): 'Already closed'
2. (auction_open = true) &&
(Amount <= Balance): 'Not higher than the current highest bid'
3. (auction_open = true) &&
(Amount <= Balance) &&
get_contract (Unit, highest_bidder) = None: 'Invalid address' 
\end{lstlisting}
